Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-08-18 manuscript/draft.Rmd
Editorial work on Intro
Includes moving Table 1 into methods where the detail present in it makes more
sense. Replaces the reference to that table in the intro with a reference to
Figure 1, which works better there.

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,200 of sites located throughout North America, while following best
practices for ecological forecasting (Box 1). To do this, we combine 32 years of
time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. Datasets that
span a time scale of 30 years or more have only recently become available for
large-scale time-series based forecasting. A dataset of this size allows us to
model and assess changes a decade or more into the future in the presence of
shifts in environmental conditions on par with predicted climate change. We
compare traditional distribution modeling based approaches to spatial models of
species richness, time-series methods, and two simple baselines that predict
constant richness for each site, on average (Figure 1). All of our forecasting
models account for uncertainty and observation error, are evaluated across
different time lags using hindcasting, and are publicly archived to allow future
assessment. We discuss the implications of these practices for our understanding
of, and confidence in, the resulting forecasts, and how we can continue to build
on these approaches to improve ecological forecasting in the future.

@@

We evaluated 6 types of forecasting models (Table 1) at two time-scales. In the
first one, we divided our 32 years of data into 22 years of training data
and retained 10 years of data for evaluation via hindcasting. Second, we used
the full data set for training and made predictions through the year 2050. For
both time scales, we evaluated versions of each model with and without
correcting for observer effects, as described below.

@@

For each forecasting approach, we trained two versions of the corresponding
model: one with corrections for differences among observers, and one without.
Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. We estimated the observer
effects (and associated uncertainty about those effects) with a linear mixed
model with observer as a random effect built in the Stan probabilistic
programming language [@stan]. Because observer and site are strongly related
(observers tend to repeatedly sample the same site), site was also included as a
random effect to ensure that inferred deviations were actually observer-related
(as opposed to being related to the sites that a given observer happened to
see). The resulting model partitions the variance in observed richness values
into site-level variance, observer-level variance, and residual variance (e.g.
variation within a site from year to year; Figure 3). The site-level estimates
can also be used directly as the "average" baseline model (see below). The
estimated observer effects can be subtracted from the richness values for a
particular observer to provide an estimate of how many species would have been
found by a "typical" observer. To incorporate uncertainty in these "corrected"
richness values into the forecasting models we collected 500 Monte Carlo samples
from the model's posterior distribution, and fit each of the downstream models
with each of the Monte Carlo samples. Each Monte Carlo sample represented a
different possible set of observer-level and site-level random effect values
across the full 32-year dataset.

>>>>

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,200 of sites located throughout North America, while following best
practices for ecological forecasting (Box 1). To do this, we combine 30 years of
time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. Datasets that
span a time scale of 30 years or more have only recently become available for
large-scale time-series based forecasting. A dataset of this size allows us to
model and assess changes a decade or more into the future in the presence of
shifts in environmental conditions on par with predicted climate change. We
compare traditional distribution modeling based approaches to spatial models of
species richness, time-series methods, and two simple baselines that predict
constant richness for each site, on average (Figure 1). All of our forecasting
models account for uncertainty and observation error, are evaluated across
different time lags using hindcasting, and are publicly archived to allow future
assessment. We discuss the implications of these practices for our understanding
of, and confidence in, the resulting forecasts, and how we can continue to build
on these approaches to improve ecological forecasting in the future.

@@

We evaluated 6 types of forecasting models (Table 1) at two time-scales. In the
first one, we divided our [30] years of data into [22] years of training data
and retained [9] years of data for evaluation via hindcasting. Second, we used
the full data set for training and made predictions through the year 2050. For
both time scales, we evaluated versions of each model with and without
correcting for observer effects, as described below.

@@

For each forecasting approach, we trained two versions of the corresponding
model: one with corrections for differences among observers, and one without.
Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. We estimated the observer
effects (and associated uncertainty about those effects) with a linear mixed
model with observer as a random effect built in the Stan probabilistic
programming language [@stan]. Because observer and site are strongly related
(observers tend to repeatedly sample the same site), site was also included as a
random effect to ensure that inferred deviations were actually observer-related
(as opposed to being related to the sites that a given observer happened to
see). The resulting model partitions the variance in observed richness values
into site-level variance, observer-level variance, and residual variance (e.g.
variation within a site from year to year; Figure 3). The site-level estimates
can also be used directly as the "average" baseline model (see below). The
estimated observer effects can be subtracted from the richness values for a
particular observer to provide an estimate of how many species would have been
found by a "typical" observer. To incorporate uncertainty in these "corrected"
richness values into the forecasting models we collected 500 Monte Carlo samples
from the model's posterior distribution, and fit each of the downstream models
with each of the Monte Carlo samples. Each Monte Carlo sample represented a
different possible set of observer-level and site-level random effect values
across the full 30-year dataset.
